---
title: "Automate Microbiome Analysis Using Quarto"
author: "Sudarshan A. Shetty"
format: html
editor: visual
---

::: callout-note
Before you start, please familiarize yourself with [RStudio](https://education.rstudio.com/learn/beginner/) and its [RStudio projects](https://www.r-bloggers.com/2020/01/rstudio-projects-and-working-directories-a-beginners-guide/).
:::

As someone analyzing microbiome data, you may find yourself frequently working with different data sets and performing the same basic analyses over and over again. These standard tasks, such as quality checking, calculating alpha and beta diversity, and visualizing the composition between groups, can be time-consuming if you're constantly copy-pasting code or searching for solutions on Stack Overflow or GitHub.

Wouldn't it be great if you could automate these analyses? That's where [Quarto](https://quarto.org/) comes in. Developed by [Posit](https://posit.co/) (formerly RStudio), [Quarto]{style="color:#dd1c77"} is the next generation version of [Rmarkdown]{style="color:#dd1c77"} (`.rmd`) tool that makes it easy to perform these common tasks and more. In fact, if you're already familiar with [Rmarkdown]{style="color:#dd1c77"}, you'll find the [Quarto]{style="color:#dd1c77"} document (`.qmd`) very similar  to `.rmd`. [Here](https://quarto.org/docs/get-started/index.html) you can find the installation details.

Here's a simple example of what you can do with [Quarto]{style="color:#dd1c77"}: read in a phyloseq object stored as an \*.rds file, calculate reads per sample, Shannon diversity, and Observed ASVs, and investigate whether there is a correlation between reads per sample and Shannon diversity, Observed ASVs. I use the data from [Fuentes et al.](https://www.thelancet.com/journals/lanhl/article/PIIS2666-7568(20)30034-9/fulltext) that consists of gut microbiota profiles from older adults with (two time points, L1 and L2) and without (C) influenza-like illness. This information is stored in the column called [ILI]{style="color:#dd1c77"} in the `sample_data` of the [phyloseq](https://joey711.github.io/phyloseq/) object `FuentesIliGutData.rds`. You can access the data from the [GitHub repo](https://github.com/microsud/quarto_microbiome/tree/main/data) of this website.  

{{< include simple_intro/simple_intro >}}

As you may notice this is one of the common tasks in a microbiota profiling study. This can be boring if you're constantly writing or copying the same code.

But what if you could automate these tasks and save yourself time and effort? In a recent post, Mike Mahoney demonstrated [How to use Quarto for Parameterized Reporting](https://www.mm218.dev/posts/2022-08-04-how-to-use-quarto-for-parameterized-reporting/). This inspired me to try [Quarto]{style="color:#dd1c77"} for automating my analysis.

One of the handy features of [Quarto]{style="color:#dd1c77"} is the [params]{style="color:#3182bd"} option, which can be used in the `.qmd` file to customize your analysis for each project. This option is also available in the [Rmarkdown]{style="color:#dd1c77"} format. Here's an example of how to use [params]{style="color:#3182bd"} in `.qmd`:

{{< include params_example/params_example >}}

Did you notice the [pseq_path: "data/FuentesIliGutData.rds"]{style="color:#3182bd"} and [primary_var: ILI]{style="color:#3182bd"} variables in the [params]{style="color:#3182bd"} section of `.qmd` file? These are [floating]{style="color:#3182bd"} variables that can be customized for each project. For every new data set, you can change the `pseq_path` and `primary_var` and render the `.qmd` file.  

To make it even easier to reuse this analysis, you can save the file for example [basic_analysis.qmd]{style="color:#dd1c77"} in your [RProject]{style="color:#dd1c77"} directory and write a simple function called `build_my_report()` that processes the `.qmd` file and renders it as an `html` report.    

Here's a draft of what the full [basic_analysis.qmd]{style="color:#dd1c77"} might look like. 

{{< include params_example_full/params_example_full >}}

Notice that the [ILI]{style="color:#dd1c77"} variable has been replaced with `primary_var` in the `params` and in the plotting `aes_string`. This allows you to customize the analysis for each project simply by changing the values of these floating variables in your `build_my_report()` function.   

Below is the `build_my_report()` function.  
```{r eval=FALSE}
build_my_report <- function(qmd_file = NULL,
                            pseq_path = NULL,
                            primary_var = NULL) {
    
    {
        # extracting the name of the source .qmd file
        base_name <- qmd_file
        base_name <- gsub(".qmd", "", base_name)
        # render the file 
        quarto::quarto_render(input = qmd_file,
                              # We add the suffix '_report' to output html file
                              output_file = paste0(base_name, "_report.html"),
                              quiet = FALSE,
                              # supply the params arguments 
                              execute_params = list("pseq_path" = pseq_path,
                                                    "primary_var" = primary_var),
                              execute_debug = FALSE,
                              cache = NULL,
                              cache_refresh = FALSE,
                              debug = FALSE)
    }
}
```

The `build_my_report(qmd_file, pseq_path, primary_var)` function takes three arguments:

-   [qmd_file]{style="color:#c51b8a"} is the name of the `.qmd` template file (e.g. [basic_analysis.qmd]{style="color:#dd1c77"}).

-   [pseq_path]{style="color:#c51b8a"} is the path to the phyloseq object that you want to analyze.

-   [primary_var]{style="color:#c51b8a"} is the name of the primary variable you want to investigate. In this case, [ILI]{style="color:#dd1c77"} is the column name in the `sample_data()` of the `FuentesIliGutData` phyloseq object.  

You can create an `RProject` and organize your files. Here is one simple example.  
``` markdown
Structure of your `RProject`:   
my_project/   
    |----- my_project.Rproj # short description of what it contains.
    |----- basic_analysis.qmd # Template .qmd file.  
    |----- build_my_report.R # R file with function to render report from .qmd file. 
    |----- data/ # A folder where data is stored.  
            |----- FuentesIliGutData.rds
```

After this you can simply `source` the function from the file `build_my_report.R` in your session.   

```{r eval=FALSE}
source("build_my_report.R") # source function to run
```

And run `build_my_report()` by supplying the necessary files/info to the arguments.    
```{r eval=FALSE}
# run function 
build_my_report(qmd_file = "basic_analysis.qmd",
                pseq_path = "data/FuentesIliGutData.rds",
                primary_var = "ILI")  
```

::: callout-note
Check the rendered html file in the tab **Example Report** above!
:::

I think that [Quarto]{style="color:#dd1c77"} is a useful tool for microbiome researchers looking to streamline their workflows and save time on routine analyses. By using the [params]{style="color:#3182bd"} option in the `.qmd` file, you can easily customize your analysis for each project, while the ability to store your `.qmd` file in your `RProject` directory means you can easily reuse your analysis across multiple projects. The reports can be standardized and shared easily with collaborators. There are several aesthetic and practical settings like hiding codes, defining visualization themes, etc as well as using the `knitr::opts_chunk` for setting global rules for your documents. Moreover, beyond `html`, reports can be generated as `pdf` or `word doc` files. You can familiarize yourself with [Quarto quide](https://quarto.org/docs/guide/) and explore the numerous options for automating your data analysis workflows. Give it a try and see how it can change the way you approach your routine microbiome data analytics.

------------------------------------------------------------------------

© Copyright 2023 Sudarshan A. Shetty
